---
title: "Data Camp Open Vaccine: Data preprocessing and visualisation"
author: "AKROUT Leyth, BOULAHFA Jawad, DE SANTIAGO Kylliann"
date: "February 6th, 2021"
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
    df_print : paged
---

```{r, echo=FALSE}
rm(list = ls())
```

```{r, results='hide', message=FALSE, warning=FALSE, echo=FALSE}
## Packages

library(tidyverse)
library(reshape2)
library(data.table)

library(gtools)

library(forcats)

library(stringr)
library(stringi)
library(readr)

library(caret)
```

# Data preprocessing

```{r, warning=FALSE, message=FALSE}
# We start by loading the data
data_filtered <- read_csv("data_filtered.csv")
# We add column names to the first two columns
colnames(data_filtered)[1] <- "Position"
colnames(data_filtered)[2] <- "Covariable"
```


```{r, warning=FALSE, message=FALSE}
# We create a list which contains the names of the nucleotides:
# 0 = A, 1 = U, 2 = C, 3 = G
nucleotide_labels <- c("A", "U", "C", "G")
#↨Labeller for facet_wrap
nucleotide_labeller <- as_labeller(c("0" = "A", "1" = "U", "2" = "C", "3" = "G"))

# We create a list which contains the names of the predicted loop types:
# 0 = S, 1 = M, 2 = I, 3 = B, 4 = H, 5 = E, 6 = X
predicted_loop_type_labels <- c("S", "M", "I", "B", "H", "E", "X")
#↨Labeller for facet_wrap
predicted_loop_type_labeller <- as_labeller(c("0" = "S", "1" = "M", "2" = "I",
                                              "3" = "B", "4" = "H", "5" = "E", "6" = "X"))

# Removing the accent
data_filtered$Covariable[data_filtered$Covariable == "nucléotide"] <- "nucleotide"

# Recoding the factor
data_filtered$Covariable <-
  recode_factor(data_filtered$Covariable, "score_structure_boucle" = "reactivity")
```

```{r}
# As "Position" is the column name, we can just extract the number inside the strings to keep the position information.
data_filtered <- data_filtered %>%
  mutate(Position = as.factor(parse_number(as.character(Position))))
```

```{r}
# First 25 lines of the data frame
head(data_filtered, n = 25)
```

```{r}
# We gather the nucleotides values for all positions in a dataframe.
nucleotide_df <- data_filtered %>% subset(Covariable == "nucleotide") %>%
  dplyr::select(-c(Position, Covariable)) %>% mutate_all(as.factor)
# The row index changed, so we change it back.
# The row index is now the position in the sequence.
rownames(nucleotide_df) <- NULL

# X is a vector which will contain all nucleotides values (for each position and each individual)
X <- c()
for (i in seq(1:nrow(nucleotide_df))){
  X <- c(X, nucleotide_df[i, ])
}
X <- unlist(X)

# We gather the predicted loop types values for all positions in a dataframe.
predicted_loop_type_df <- data_filtered %>% subset(Covariable == "structure_boucle") %>%
  dplyr::select(-c(Position, Covariable))
# The row index changed, so we change it back.
# The row index is now the position in the sequence.
rownames(predicted_loop_type_df) <- NULL

predicted_loop_type_vector <- c()
for (i in seq(1:nrow(predicted_loop_type_df))){
  predicted_loop_type_vector <- c(predicted_loop_type_vector, predicted_loop_type_df[i, ])
}
predicted_loop_type_vector <- unlist(predicted_loop_type_vector)

# We gather the reactivity values for all positions in a dataframe.
reactivity_df <- data_filtered %>%
  subset(Covariable == "reactivity") %>%
  dplyr::select(-c(Position, Covariable))
rownames(reactivity_df) <- NULL

# We gather the score_mg_ph10 values for all positions in a dataframe.
score_mg_ph10_df <- data_filtered %>%
  subset(Covariable == "score_mg_ph10") %>%
  dplyr::select(-c(Position, Covariable))
# The row index changed, so we change it back.
# The row index is now the position in the sequence.
rownames(score_mg_ph10_df) <- NULL

# We gather the score_ph10 values for all positions in a dataframe.
score_ph10_df <- data_filtered %>%
  subset(Covariable == "score_ph10") %>%
  dplyr::select(-c(Position, Covariable))
# The row index changed, so we change it back.
# The row index is now the position in the sequence.
rownames(score_ph10_df) <- NULL

# We gather the score_mg_50c values for all positions in a dataframe.
score_mg_50c_df <- (data_filtered[data_filtered[,2]=='score_mg_50c',]) %>% dplyr::select(-c(Position, Covariable))
# The row index changed, so we change it back.
# The row index is now the position in the sequence.
rownames(score_mg_50c_df) <- NULL

# We gather the score_50c values for all positions in a dataframe.
score_50c_df <- data_filtered %>%
  subset(Covariable == "score_50c") %>%
  dplyr::select(-c(Position, Covariable))
# The row index changed, so we change it back.
# The row index is now the position in the sequence.
rownames(score_50c_df) <- NULL

# Y_reactivity, ... ,Y_score_50c are vector which will contain all score values (for each position and each individual)
Y_reactivity <- c()
Y_score_mg_ph10 <- c()
Y_score_ph10 <- c()
Y_score_mg_50c <- c()
Y_score_50c <- c()
for (i in seq(1:nrow(nucleotide_df)))
{
  Y_reactivity <- c(Y_reactivity, reactivity_df[i, ])
  Y_score_mg_ph10 <- c(Y_score_mg_ph10, score_mg_ph10_df[i, ])
  Y_score_ph10 <- c(Y_score_ph10, score_ph10_df[i, ])
  Y_score_mg_50c <- c(Y_score_mg_50c, score_mg_50c_df[i, ])
  Y_score_50c <- c(Y_score_50c, score_50c_df[i, ])
} 
Y_reactivity <- unlist(Y_reactivity)
Y_score_mg_ph10 <- unlist(Y_score_mg_ph10)
Y_score_ph10 <- unlist(Y_score_ph10)
Y_score_mg_50c <- unlist(Y_score_mg_50c)
Y_score_50c <- unlist(Y_score_50c)

scores_df <- data.frame(X, Y_reactivity, Y_score_mg_ph10,
                        Y_score_ph10, Y_score_mg_50c, Y_score_50c)
```


# Data visualisation

## Nucleotides

### Pie plot

```{r}
nucleotide_table <- data.frame(table(X)) %>% mutate(Proportion = Freq/sum(Freq)) %>%
  mutate(Percent = 100*Proportion) %>% mutate(Nucleotide = nucleotide_labels) %>%
  dplyr::select(X, Nucleotide, everything())
colnames(nucleotide_table)[3] <- "Count" 
```

```{r}
nucleotide_table
```

There are 108 052 nucleotides in the dataset.
```{r}
nucleotide_table$Count %>% sum()
```

```{r}
nucleotide_table %>%
  mutate(ypos = cumsum(Proportion) - 0.5 * Proportion) %>%
  ggplot(aes(x="", y=Proportion, fill=Nucleotide)) +
  geom_bar(stat="identity", width=1, color="black") +
  coord_polar(theta = "y", direction = 1) + theme_void() +
  geom_text(aes(label = paste0(round(100*Proportion, 2), "%")),
            position = position_stack(vjust = 0.5),
            color = "white", size = 5) +
  ggtitle("Nucleotides pie plot")
```

### Boxplots

```{r}
scores_df %>% ggplot(aes(x = X, y = Y_reactivity)) + geom_boxplot(aes(color = X)) + coord_flip() +
  labs(x = "nucléotides", y = "reactivity", color = "nucléotides") +
  scale_x_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```


```{r}
scores_df %>% ggplot(aes(x = X, y = Y_score_mg_ph10)) +
  geom_boxplot(aes(color = X)) + coord_flip() +
  labs(x = "nucléotides", y = "score_mg_ph10", color = "nucléotides") +
  scale_x_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```


```{r}
scores_df %>% ggplot(aes(x = X, y = Y_score_ph10)) +
  geom_boxplot(aes(color = X)) + coord_flip() +
  labs(x = "nucléotides", y = "score_ph10", color = "nucléotides") +
  scale_x_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```


```{r}
scores_df %>% ggplot(aes(x = X, y = Y_score_mg_50c)) +
  geom_boxplot(aes(color = X)) + coord_flip() +
  labs(x = "nucléotides", y = "score_mg_50c", color = "nucléotides") +
  scale_x_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```


```{r}
scores_df %>% ggplot(aes(x = X, y = Y_score_50c)) +
  geom_boxplot(aes(color = X)) + coord_flip() +
  labs(x = "nucléotides", y ="score_50c", color = "nucléotides") +
  scale_x_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```

## Predicted loop types

### Pie plot

```{r}
predicted_loop_type_table <- data.frame(table(predicted_loop_type_vector)) %>%
  mutate(Proportion = Freq/sum(Freq)) %>%
  arrange(predicted_loop_type_vector) %>%
  mutate(Percent = 100*Proportion) %>%
  mutate(Predicted_loop_type = predicted_loop_type_labels) %>%
  dplyr::select(predicted_loop_type_vector, Predicted_loop_type, everything())
colnames(predicted_loop_type_table)[3] <- "Count"
```

```{r}
predicted_loop_type_table
```

```{r}
predicted_loop_type_table$Count %>% sum()
```

```{r, fig.height=8, fig.width=10}
predicted_loop_type_table %>%
  mutate(ypos = cumsum(Proportion) - 0.5 * Proportion) %>%
  ggplot(aes(x="", y=Proportion, fill=Predicted_loop_type)) +
  geom_bar(stat="identity", width=1, color="black") +
  coord_polar(theta = "y", direction = 1, start = 0) + theme_void() +
  geom_text(aes(x = 1.63, label = paste0(round(100*Proportion, 2), "%")),
            position = position_stack(vjust = 0.5),
            color = "black", size = 4.5) +
  ggtitle("Predicted loop type pie plot")
```

### Boxplots

## Histograms and density functions

### Reactivity

```{r}
# Density function without grouping.
scores_df %>% ggplot(aes(x = Y_reactivity)) +
  geom_histogram(aes(y = ..density..), color = "black", fill = "red", alpha = 0.1) +
  geom_density(color = "red", size = 1)
```

```{r}
# Density functions according to the nucleotides.
scores_df %>% ggplot(aes(x = Y_reactivity, fill = X)) + 
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = c('#999999','#E69F00','#FC4D27','#2773FC'),
                      labels = nucleotide_labels) + 
  theme(legend.position ="right")+labs(title="reactivity", fill="nucleotide")
```

```{r}
# We use facet_wrap to have a better look on each function.
scores_df %>% ggplot(aes(x = Y_reactivity, group = X, fill = X)) + 
  geom_density() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+labs(title="reactivity", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
# We use facet_wrap to have a better look on each function.
scores_df %>% ggplot(aes(x = Y_reactivity, group = X, colour = X)) + 
  geom_histogram(aes(y = ..density.., fill = X), alpha = 0.1) +
  geom_density(size = 1) + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+
  labs(title="reactivity", color = "nucleotide", fill = "nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_mg_ph10

```{r}
# Density function without grouping.
scores_df %>% ggplot(aes(Y_score_mg_ph10)) +
  geom_histogram(aes(y = ..density..), color = "black", fill = "blue", alpha = 0.1) +
  geom_density(color = "blue", size = 1)
```

```{r}
scores_df %>% ggplot(aes(Y_score_mg_ph10, fill = X)) + 
  geom_density(alpha=.5) + 
  scale_fill_manual(values = c('#999999','#E69F00','#FC4D27','#2773FC')) + 
  theme(legend.position ="right")+labs(title="score_mg_ph10", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(Y_score_mg_ph10, group = X, fill = X)) + 
  geom_density() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+labs(title="score_mg_ph10", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
# We use facet_wrap to have a better look on each function.
scores_df %>% ggplot(aes(x = Y_score_mg_ph10, group = X, colour = X)) + 
  geom_histogram(aes(y = ..density.., fill = X), alpha = 0.1) +
  geom_density(size = 1) + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+
  labs(title="score_mg_ph10", color = "nucleotide", fill = "nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```




### score_ph10

```{r}
# Density function without grouping.
scores_df %>% ggplot(aes(Y_score_ph10)) +
  geom_histogram(aes(y = ..density..), color = "black", fill = "orange", alpha = 0.1) +
  geom_density(color = "darkorange", size = 1)
```

```{r}
scores_df %>% ggplot(aes(Y_score_ph10, fill = X)) + 
  geom_density(alpha=.5) + 
  scale_fill_manual(values = c('#999999','#E69F00','#FC4D27','#2773FC')) + 
  theme(legend.position ="right")+labs(title="score_ph10", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(Y_score_ph10, group = X, fill = X)) + 
  geom_density() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+labs(title="score_ph10", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
# We use facet_wrap to have a better look on each function.
scores_df %>% ggplot(aes(x = Y_score_ph10, group = X, colour = X)) + 
  geom_histogram(aes(y = ..density.., fill = X), alpha = 0.1) +
  geom_density(size = 1) + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+
  labs(title="score_ph10", color = "nucleotide", fill = "nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_mg_50c

```{r}
scores_df %>% ggplot(aes(Y_score_mg_50c)) +
  geom_histogram(aes(y = ..density..), color = "black", fill = "purple", alpha = 0.1) +
  geom_density(color = "purple", size = 1)
```

```{r}
scores_df %>% ggplot(aes(Y_score_mg_50c, fill=as.factor(X))) + 
  geom_density(alpha=.5) + 
  scale_fill_manual(values = c('#999999','#E69F00','#FC4D27','#2773FC')) + 
  theme(legend.position ="right")+labs(title="score_mg_50c", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(Y_score_mg_50c, group = X, fill = X)) + 
  geom_density() + facet_wrap(~X, labeller = nucleotide_labeller) +
  theme(legend.position ="right")+labs(title="score_mg_50c", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
# We use facet_wrap to have a better look on each function.
scores_df %>% ggplot(aes(x = Y_score_mg_50c, group = X, colour = X)) + 
  geom_histogram(aes(y = ..density.., fill = X), alpha = 0.1) +
  geom_density(size = 1) + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+
  labs(title="score_mg_50c", color = "nucleotide", fill = "nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```




### score_50c

```{r}
scores_df %>% ggplot(aes(Y_score_50c)) +
  geom_histogram(aes(y = ..density..), color = "black", fill = "brown", alpha = 0.1) +
  geom_density(color = "brown", size = 1)
```

```{r}
scores_df %>% ggplot(aes(Y_score_50c, fill = X)) + 
  geom_density(alpha=.5) + 
  #scale_fill_manual(values = c('#999999','#E69F00','#FC4D27','#2773FC')) + 
  theme(legend.position ="right")+labs(title="score_50c", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(Y_score_50c, group = X, fill = X)) + 
  geom_density() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+labs(title="score_50c", fill="nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels)
```

```{r}
# We use facet_wrap to have a better look on each function.
scores_df %>% ggplot(aes(x = Y_score_50c, group = X, colour = X)) + 
  geom_histogram(aes(y = ..density.., fill = X), alpha = 0.1) +
  geom_density(size = 1) + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right")+
  labs(title="score_50c", color = "nucleotide", fill = "nucleotide") +
  scale_fill_discrete(labels = nucleotide_labels) +
  scale_color_discrete(labels = nucleotide_labels)
```

## Scatter plots

### score_mg_50c depending on score_50c

```{r}
scores_df %>% ggplot(aes(x = Y_score_50c, y = Y_score_mg_50c, color = X)) + 
  geom_point() +
  theme(legend.position ="right") +
  labs(x = "score_50c", y = "score_mg_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(x = Y_score_50c, y = Y_score_mg_50c, group = X, color = X)) + 
  geom_point() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right") +
  labs(x = "score_50c", y = "score_mg_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_mg_50c depending on score_mg_ph10

```{r}
scores_df %>% ggplot(aes(x = Y_score_mg_ph10, y = Y_score_mg_50c, color = X)) + 
  geom_point() +
  theme(legend.position ="right") +
  labs(x="score_mg_ph10", y = "score_mg_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(x = Y_score_mg_ph10, y = Y_score_mg_50c, group = X, color = X)) + 
  geom_point() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right") +
  labs(x="score_mg_ph10", y = "score_mg_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_mg_50c depending on score_ph10

```{r}
scores_df %>% ggplot(aes(x = Y_score_ph10, y = Y_score_mg_50c, color = X)) + 
  geom_point() +
  theme(legend.position ="right") +
  labs(x="score_ph10", y = "score_mg_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(x = Y_score_ph10, y = Y_score_mg_50c, group = X, color = X)) + 
  geom_point() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right") +
  labs(x="score_ph10", y = "score_mg_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_mg_ph10 depending on score_ph10

```{r}
scores_df %>% ggplot(aes(x = Y_score_ph10, y = Y_score_mg_ph10, color = X)) + 
  geom_point() +
  theme(legend.position ="right") +
  labs(x = "score_ph10", y = "score_mg_ph10", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(x = Y_score_ph10, y = Y_score_mg_ph10, group = X, color = X)) + 
  geom_point() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right") +
  labs(x = "score_ph10", y = "score_mg_ph10", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_mg_ph10 depending on score_50c

```{r}
scores_df %>% ggplot(aes(x = Y_score_50c, y = Y_score_mg_ph10, color = X)) + 
  geom_point() +
  theme(legend.position ="right") +
  labs(x = "score_50c", y = "score_mg_ph10", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(x = Y_score_50c, y = Y_score_mg_ph10, group = X, color = X)) + 
  geom_point() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right") +
  labs(x = "score_50c", y = "score_mg_ph10", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

### score_50c depending on score_ph10

```{r}
scores_df %>% ggplot(aes(x = Y_score_ph10, y = Y_score_50c, color = X)) + 
  geom_point() +
  theme(legend.position ="right") +
  labs(x = "score_ph10", y = "score_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```

```{r}
scores_df %>% ggplot(aes(x = Y_score_ph10, y = Y_score_50c, group = X, color = X)) + 
  geom_point() + facet_wrap(~X, labeller = nucleotide_labeller) + 
  theme(legend.position ="right") +
  labs(x = "score_ph10", y = "score_50c", color="nucleotide") +
  scale_color_discrete(labels = nucleotide_labels)
```


